### 2020-09-01 manuscript.tex
complete draft

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended
% to minimize problems and delays during our production
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% Once your paper is accepted for publication,
% PLEASE REMOVE ALL TRACKED CHANGES in this file
% and leave only the final text of your manuscript.
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file.
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column.
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

The molecular clock hypothesis states that the evolutionary rates of biological sequences are approximately constant through time \cite{zuckerkandl1962molecular}.
This assumption forms the basis of phylogenetics, under which the evolutionary trees and divergence dates of life forms are inferred from biological sequences, such as nucleic and amino acids \cite{douzery2003local, drummond2006relaxed}.
In Bayesian phylogenetics, these trees and their associated parameters are estimated as probabilitiy distributions \cite{kuhner1995estimating, larget1999markov, mau1999bayesian}. 
Statistical inference can be performed by the Markov chain Monte Carlo (MCMC) algorithm \cite{metropolis53, hastings70} using platforms such as BEAST \cite{drummond2012bayesian}, BEAST2 \cite{bouckaert2019beast}, MrBayes \cite{ronquist2012mrbayes}, and RevBayes \cite{hohna2016revbayes}.

The simplest phylogenetic clock model -- the strick clock -- makes the mathematically convenient assumption that the evolutionary rate is constant across all lineages \cite{zuckerkandl1965evolutionary, kuhner1995estimating, larget1999markov}.  % Early examples? 
However, molecular substitution rates are known to vary over time, over population sizes, over evolutionary pressures, and over nucleic acid replicative machineries \cite{gillespie1994causes, woolfit2009effective, loh2010optimization}.
Moreover, any given dataset may be clock-like (where the substitution rates have a small variance across lineages) or non clock-like (a large variance). 
In the latter case, a strict clock is probably not suitable.

@@

With the overwhelming availabilty of biological sequence data, the development of efficient Bayesian phylogenetic methods is more important than ever. 
The performance of MCMC is dependent not only on computational runtime but also the efficacy of an
MCMC setup to achieve its convergence.
A critical task therein lies the further advancement of MCMC operators.
Recent developments in this area include the advancement of guided tree proposals \cite{zhang2020using, meyer2019adaptive, hohna2012guided}, coupled MCMC \cite{altekar2004parallel, muller2019coupled},
adaptive multivariate transition kernels \cite{baele2017adaptive},
and other explorative proposal kernels such as the Bactrian and mirror kernels \cite{yang2013searching, thawornwattana2018designing}.
In the case of clock models, informed tree proposals can account for correlations between substitution rates and divergence times \cite{zhang2020improving}. 
The rate parameterisation itself can also affect the ability to ``mix'' during MCMC \cite{drummond2006relaxed, li2012model, zhang2020improving}.

@@

The weight of an operator determines the probability of the operator being selected and is typically fixed throughout MCMC.
In BEAST2, operators can have its own tunable parameter $s$ which determines the step size of the operator, and this term is learned over the course of the MCMC \cite{bouckaert2019beast}. 
Here, we define clock model operators as those which generate proposals for either $\vec{\mathcal{R}}^{\,}$ or $\sigma$.
Pre-existing BEAST2 clock model operators are summarised in \textbf{Table \ref{table:kernels}}, and further operators are introduced throughout the paper.

\begin{table}[h!]
\centering
\begin{tabular}{l p{4.2cm} l l} 
 \hline
 Operator & Description & Parameters & Parameterisations  \\
  \hline
 \texttt{RandomWalk} & Moves a single element by a tunable amount. & $\vec{\mathcal{R}}^{\,}, \sigma$ & \textit{cat}, \textit{real}, \textit{quant} \\
  \hline
\texttt{Scale} & Applies \texttt{RandomWalk} on the log-transformation (suitable for parameters with positive domains). & $\vec{\mathcal{R}}^{\,}, \sigma$ & \textit{real}, \textit{quant}  \\
  \hline
 \texttt{Interval} & Applies \texttt{RandomWalk} on the logit-transformation (suitable for parameters with upper and lower limits). & $\vec{\mathcal{R}}^{\,}$ & \textit{quant}  \\
  \hline
 \texttt{Swap} & Swaps two random elements in the vector \cite{drummond2006relaxed}. & $\vec{\mathcal{R}}^{\,}$  & \textit{cat}, \textit{real}, \textit{quant}  \\
 \hline
\texttt{Uniform} & Resamples one element in the vector from a uniform distribution. & $\vec{\mathcal{R}}^{\,}$  & \textit{cat}, \textit{quant}  \\
 \hline
\texttt{ConstantDistance} & Adjusts an internal node height and recalculates all incident branch rates such that the genetic distances remain constant \cite{zhang2020improving}.  & $\vec{\mathcal{R}}^{\,}, \mathcal{T}$ & \textit{real}, \textit{quant} \\
 \hline
\texttt{SimpleDistance} & Applies \texttt{ConstantDistance} to the root node \cite{zhang2020improving}.  & $\vec{\mathcal{R}}^{\,}, \mathcal{T}$ & \textit{real}, \textit{quant} \\
 \hline
\texttt{SmallPulley} & Proposes new branch rates incident to the root such that their combined genetic distance is constant  \cite{zhang2020improving}.  & $\vec{\mathcal{R}}^{\,}$ & \textit{real}, \textit{quant} \\
 \hline
\texttt{CisScale} & Applies \texttt{Scale} to $\sigma$. Then recomputes all rates such that their quantiles are constant (for \textit{real} \cite{zhang2020improving}) or recomputes all quantiles such that their rates are constant (\textit{quant}).  & $\vec{\mathcal{R}}^{\,}, \sigma$ & \textit{real}, \textit{quant} \\
 \hline
\end{tabular}
\caption{Summary of pre-existing BEAST2 operators, which apply to either branch rates $\vec{\mathcal{R}}^{\,}$ or the clock standard deviation $\sigma$, and the substitution rate parameterisation they apply to.
 \texttt{ConstantDistance} and \texttt{SimpleDistance} also adjust node heights in the tree $\mathcal{T}$. }
\label{table:kernels}
\end{table}

The family of constant distance operators (\texttt{ConstantDistance}, \texttt{SimpleDistance}, and \texttt{SmallPulley} \cite{zhang2020improving}) are best suited for larger datasets (or datasets with strong signal) where the likelihood distribution is peaked (\textbf{Fig. \ref{fig:rateparams}}).
While simple one dimensional operators such as \texttt{RandomWalk} or \texttt{Scale} must make small steps in order to stay ``on the ridge'' of the likelihood function, the constant distance operators ``wander along the ridge''  by ensuring that genetic distances are constant after the proposal.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/correlations.pdf}
\caption{\textbf{Traversing likelihood space.}
The z-axes above are the log-likelihoods of the genetic distance $r \times \tau$ between two simulated nucleic acid sequences of length $L$, under the Jukes-Cantor substitution model \cite{jukes1969evolution}. 
Two possible proposals from the current state (white circle) are depicted.
These proposals are generated by the \texttt{RandomWalk} (\texttt{RW}) and \texttt{ConstantDistance} (\texttt{CD}) operators.
In the low signal dataset ($L=0.1$kb), both operators can traverse the likelihood space effectively.
 However, the exact same proposal by \texttt{RandomWalk} incurs a much larger likelihood penalty in the $L=0.5$kb dataset by ``falling off the ridge'', in contrast to \texttt{ConstantDistance} which ``walks along the ridge''.
 This discrepancy is even stronger for larger datasets and thus necessitates the use of operators such as \texttt{ConstantDistance} which account for correlations between branch lengths and rates. }
\label{fig:rateparams}
\end{figure}

 \texttt{Scale} and \texttt{CisScale} both operate on the clock model standard deviation $\sigma$ however they behave differently in the \textit{real} and \textit{quant} parameterisations (\textbf{Fig. \ref{fig:clockSDoperators}}).
In \textit{real}, large proposals of $\sigma \rightarrow \sigma^\prime$ made by  \texttt{Scale} are likely to be rejected due to large penalties in clock model prior $p(\vec{\mathcal{R}}^{\,}|\sigma^\prime)$.
This led to the development of the fast clock scaler \cite{zhang2020improving} (herein referred to as \texttt{CisScale}), which recomputes all branch rates $\vec{\mathcal{R}}^{\,} \rightarrow \vec{\mathcal{R}}^{\, \prime}$ such that their quantiles under the new clock model prior remain constant $p(\vec{\mathcal{R}}^{\,}|\sigma) = p(\vec{\mathcal{R}}^{\, \prime}|\sigma^\prime)$.
In contrast, a proposal made by \texttt{Scale} $\sigma \rightarrow \sigma^\prime$ under the \textit{quant} parameterisation implicitly alters all branch rates $r(\vec{\mathcal{R}}^{\,})$ while leaving the quantiles $\vec{\mathcal{R}}^{\,}$ themselves constant.
But by applying \texttt{CisScale} under \textit{quant},
all quantiles are recomputed $\vec{\mathcal{R}} \rightarrow \vec{\mathcal{R}}^\prime$  such that the respective branch rates are constant i.e. $r(\vec{\mathcal{R}}^{\,}) = r(\vec{\mathcal{R}}^{\,\prime})$.
Overall, \texttt{Scale} and \texttt{CisScale} propose rates/quantiles in the opposite (trans) or same (cis) space that the clock model is parameterised under (\textbf{Fig. \ref{fig:clockSDoperators}}).

@@

where \texttt{RF} is the Robinson-Foulds tree distance \cite{robinson1981comparison}, and $|p|$ is the number of dimensions of numerical parameter $p$ (1 for $\sigma$, $2N-2$ for $\vec{\mathcal{R}}^{\,}$, and $2N-1$ for node heights $t$).
The remaining terms are trained during the second and third phases: the cumulative computational runtime spent on each operator $\mathbb{T}(\omega_i)$, the sample standard deviation $\sigma_p$ of each numerical parameter $p$, and the sum distances $\sum_x \mathbb{D}(x_p, x_p^\prime)$.

@@

Datasets with strong signal (or large $L$) are likely to achieve faster convergence when the more precise and meticulous kinds of operators are employed, such as those informed by correlations within the posterior distribution (\textbf{Fig. \ref{fig:rateparams}}).
Whereas, datasets which contain very poor signal (or small $L$) are likely to mix faster when more weight is placed on bold operators. 
We therefore introduce the \texttt{SampleFromPrior($\vec{x}^{\,}$)} operator.
This operator resamples $\psi$ randomly selected elements within vector $\vec{x}^{\,}$ from their prior distributions, where $\psi \sim \text{Binomial}(n=|\vec{x}^{\,}|, p=\frac{s}{|\vec{x}^{\,}|})$ for tunable term $s$.
%The \texttt{Uniform} operator is a special case of \texttt{SampleFromPrior}.
\texttt{SampleFromPrior} is always included among the mixture of operators under \texttt{AdaptiveOperatorSampler} and serves to make the boldest proposals for datasets with poor signal.
Ideally, an operator to the likes of \texttt{AdaptiveOperatorSampler} would learn the combination of weights behind these classes of operators best suited for any given dataset.

@@

The step size of a proposal kernal $q(x^\prime|x)$ should be such that the proposed state $x^\prime$ is sufficiently far from the current state $x$ to explore vast areas of parameter space, but not so large that the proposal is rejected too often \cite{roberts1997weak}. 
Operators which attain an acceptance probability of 0.234 are often considered to have arrived at a suitable midpoint between these two extremes \cite{bouckaert2019beast, roberts1997weak}.
The standard uniform distribution kernel has recently been challenged by the Bactrian kernel \cite{yang2013searching, thawornwattana2018designing}.
The $\text{Bactrian}(m)$ distribution is defined as the sum of two Normal distributions:

@@

where $0 \leq m < 1$ describes the modality of the Bactrian distribution. When $m=0$, the Bactrian distribution is equivalent to a $\text{Normal}(0, 1)$ distribution. As $m \rightarrow 1$, the distribution becomes increasingly bimodal (\textbf{Fig. \ref{fig:bactrian}}). 
Yang et al. 2013 \cite{yang2013searching} suggest that $\text{Bactrian}(m=0.95)$ yields a proposal kernel which traverses the posterior distribution more efficiently that the standard uniform kernel, by placing minimal probability on steps which are too small or too large.
In this case, a target acceptance probability of around 0.3 is optimal.

@@

% Give the equation of the Uniform + Bactrian kernals
% Fig: Bactrian m=0, 0.92, 0.95, 0.98, + Uniform(-1,1)
% In this article we compare the Uniform + Bactrian proposals in the clock model

@@

The \texttt{NarrowExchange} operator \cite{drummond2002estimating}, used widely in BEAST \cite{drummond2012bayesian,suchard2018bayesian} and BEAST2 \cite{bouckaert2019beast}, is similar to nearest-neighbour-interchange \cite{semple2003phylogenetics}, and works as follows (\textbf{Fig. \ref{fig:narrowexchange}}):

@@

Lakner et al. 2008 \cite{lakner2008efficiency} found that tree operators which perterb topology (such as nearest-neighbour-interchange and subtree-prune-and-regraft \cite{semple2003phylogenetics}) consistently perform better than those which also change branch lengths (such as LOCAL \cite{simon1998local} and Continuous Change \cite{jow2002bayesian}). 
If \texttt{NarrowExchange} was adapted the relaxed clock model by ensuring that genetic distances remain constant after the proposal (akin to the constant distance operators \cite{zhang2020improving}), then its ability to traverse the state space may improve. 
This may in turn permit proposing a new node height $t_D$ and therefore changing branch lengths.

@@

for random walk step size $s\Sigma$ where $s$ is a tunable scaler parameter and $\Sigma$ is drawn from a uniform or \textbf{\nameref{sect:randomwalks}}. NER (and NERw) are compatible with both the \textit{real} and \textit{quant} parameterisations. 
Analagous to the \texttt{ConstantDistance} operator, the proposed rates ensure that the genetic distances between nodes $A$, $B$, $C$, and $E$ are constant. 
There are six pairwise distance between these four nodes and therefore six constraints:

@@

\paragraph{2. Solving Jacobian determinants.} The determinant of the Jacobian matrix $J$ is required for computng the Green ratio of the proposal. $J$ is defined as

@@

\paragraph{3. Automated generation of BEAST2 operators.} Java class files are generated using string processing. Each class corresponds to a single operator, extends the class of a meta-NER-operator, and is comprised of the solutions found in \textbf{1} and the Jacobian determinant found in \textbf{2}. $|J|$ is further augmented if the \textit{quant} parameterisation is employed (\textbf{\nameref{sect:S1Appendix}}).
Two such operators are expressed in \textbf{Algorithms \ref{alg:NER1} and \ref{alg:NER2}}.

@@

These experiments showed that NER variants which satisfy the genetic distances between nodes $B$ and $A$ (i.e. $\mathcal{D}_{AB}$) or between $B$ and $C$ (i.e. $\mathcal{D}_{BC}$) usually performed worse than the standard \texttt{NarrowExchange} operator (\textbf{Fig. \ref{fig:acceptanceRateScreening}}).
This is an intuitive result. 
If there is high uncertainty in the positioning of $B$ with respect to $A$ and $C$, then there is no value in respecting either of these distance constraints, and the proposals made to the rates may often be too bold or the Green ratio $|J|$ too small for the proposal to be accepted.

@@

\subsection*{Round 3: Bactrian proposal kernels are 10\% more efficient than uniform kernels}

We benchmarked the adapt (\textit{real}) configuration with a) standard uniform proposal kernels, and b) Bactrian(0.95) kernels.
These kernels applied to all clock model related operators (\textbf{Table \ref{table:bactriankernels}}).

\subsection*{Round 4: Evaluating adaptive NER}

>>>>

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended
% to minimize problems and delays during our production
% process. Please follow the template instructionssum 
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% Once your paper is accepted for publication,
% PLEASE REMOVE ALL TRACKED CHANGES in this file
% and leave only the final text of your manuscript.
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file.
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column.
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

The molecular clock hypothesis states that the evolutionary rates of biological sequences are approximately constant through time \cite{zuckerkandl1962molecular}.
This assumption forms the basis of phylogenetics, under which the evolutionary trees and divergence dates of life forms are inferred from biological sequences, such as nucleic and amino acids \cite{douzery2003local, drummond2006relaxed}.
In Bayesian phylogenetics, these trees and their associated parameters are estimated as probability distributions \cite{kuhner1995estimating, larget1999markov, mau1999bayesian}. 
Statistical inference can be performed by the Markov chain Monte Carlo (MCMC) algorithm \cite{metropolis53, hastings70} using platforms such as BEAST \cite{drummond2012bayesian}, BEAST2 \cite{bouckaert2019beast}, MrBayes \cite{ronquist2012mrbayes}, and RevBayes \cite{hohna2016revbayes}.

The simplest phylogenetic clock model -- the strict clock -- makes the mathematically convenient assumption that the evolutionary rate is constant across all lineages \cite{zuckerkandl1965evolutionary, kuhner1995estimating, larget1999markov}.  % Early examples? 
However, molecular substitution rates are known to vary over time, over population sizes, over evolutionary pressures, and over nucleic acid replicative machineries \cite{gillespie1994causes, woolfit2009effective, loh2010optimization}.
Moreover, any given dataset may be clock-like (where the substitution rates have a small variance across lineages) or non clock-like (a large variance). 
In the latter case, a strict clock is probably not suitable.

@@

With the overwhelming availability of biological sequence data, the development of efficient Bayesian phylogenetic methods is more important than ever. 
The performance of MCMC is dependent not only on computational runtime but also the efficacy of an
MCMC setup to achieve its convergence.
A critical task therein lies the further advancement of MCMC operators.
Recent developments in this area include the advancement of guided tree proposals \cite{zhang2020using, meyer2019adaptive, hohna2012guided}, coupled MCMC \cite{altekar2004parallel, muller2019coupled},
adaptive multivariate transition kernels \cite{baele2017adaptive},
and other explorative proposal kernels such as the Bactrian and mirror kernels \cite{yang2013searching, thawornwattana2018designing}.
In the case of clock models, informed tree proposals can account for correlations between substitution rates and divergence times \cite{zhang2020improving}. 
The rate parameterisation itself can also affect the ability to ``mix'' during MCMC \cite{drummond2006relaxed, li2012model, zhang2020improving}.

@@

The weight of an operator determines the probability of the operator being selected and is typically fixed throughout MCMC.
In BEAST2, operators can have their own tunable parameter $s$ which determines the step size of the operator, and this term is learned over the course of the MCMC \cite{bouckaert2019beast}. 
Here, we define clock model operators as those which generate proposals for either $\vec{\mathcal{R}}^{\,}$ or $\sigma$.
Pre-existing BEAST2 clock model operators are summarised in \textbf{Table \ref{table:kernels}}, and further operators are introduced throughout the paper.

\begin{table}[h!]
\centering
\begin{tabular}{l p{4.2cm} l l} 
 Operator & Description & Parameters & Parameterisations  \\
  \hline
 \texttt{RandomWalk} & Moves a single element by a tunable amount. & $\vec{\mathcal{R}}^{\,}, \sigma$ & \textit{cat}, \textit{real}, \textit{quant} \\
  \hline
\texttt{Scale} & Applies \texttt{RandomWalk} on the log-transformation (suitable for parameters with positive domains). & $\vec{\mathcal{R}}^{\,}, \sigma$ & \textit{real}, \textit{quant}  \\
  \hline
 \texttt{Interval} & Applies \texttt{RandomWalk} on the logit-transformation (suitable for parameters with upper and lower limits). & $\vec{\mathcal{R}}^{\,}$ & \textit{quant}  \\
  \hline
 \texttt{Swap} & Swaps two random elements in the vector \cite{drummond2006relaxed}. & $\vec{\mathcal{R}}^{\,}$  & \textit{cat}, \textit{real}, \textit{quant}  \\
 \hline
\texttt{Uniform} & Resamples one element in the vector from a uniform distribution. & $\vec{\mathcal{R}}^{\,}$  & \textit{cat}, \textit{quant}  \\
 \hline
\texttt{ConstantDistance} & Adjusts an internal node height and recalculates all incident branch rates such that the genetic distances remain constant \cite{zhang2020improving}.  & $\vec{\mathcal{R}}^{\,}, \mathcal{T}$ & \textit{real}, \textit{quant} \\
 \hline
\texttt{SimpleDistance} & Applies \texttt{ConstantDistance} to the root node \cite{zhang2020improving}.  & $\vec{\mathcal{R}}^{\,}, \mathcal{T}$ & \textit{real}, \textit{quant} \\
 \hline
\texttt{SmallPulley} & Proposes new branch rates incident to the root such that their combined genetic distance is constant  \cite{zhang2020improving}.  & $\vec{\mathcal{R}}^{\,}$ & \textit{real}, \textit{quant} \\
 \hline
\texttt{CisScale} & Applies \texttt{Scale} to $\sigma$. Then recomputes all rates such that their quantiles are constant (for \textit{real} \cite{zhang2020improving}) or recomputes all quantiles such that their rates are constant (\textit{quant}).  & $\vec{\mathcal{R}}^{\,}, \sigma$ & \textit{real}, \textit{quant} \\
\end{tabular}
\caption{Summary of pre-existing BEAST2 operators, which apply to either branch rates $\vec{\mathcal{R}}^{\,}$ or the clock standard deviation $\sigma$, and the substitution rate parameterisation they apply to.
 \texttt{ConstantDistance} and \texttt{SimpleDistance} also adjust node heights in the tree $\mathcal{T}$. }
\label{table:kernels}
\end{table}

The family of constant distance operators (\texttt{ConstantDistance}, \texttt{SimpleDistance}, and \texttt{SmallPulley} \cite{zhang2020improving}) are best suited for larger datasets (or datasets with strong signal) where the likelihood distribution is peaked (\textbf{Fig \ref{fig:rateparams}}).
While simple one dimensional operators such as \texttt{RandomWalk} or \texttt{Scale} must make small steps in order to stay ``on the ridge'' of the likelihood function, the constant distance operators ``wander along the ridge''  by ensuring that genetic distances are constant after the proposal.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/correlations.pdf}
\caption{\textbf{Traversing likelihood space.}
The z-axes above are the log-likelihoods of the genetic distance $r \times \tau$ between two simulated nucleic acid sequences of length $L$, under the Jukes-Cantor substitution model \cite{jukes1969evolution}. 
Two possible proposals from the current state (white circle) are depicted.
These proposals are generated by the \texttt{RandomWalk} (\texttt{RW}) and \texttt{ConstantDistance} (\texttt{CD}) operators.
In the low signal dataset ($L=0.1$kb), both operators can traverse the likelihood space effectively.
 However, the exact same proposal by \texttt{RandomWalk} incurs a much larger likelihood penalty in the $L=0.5$kb dataset by ``falling off the ridge'', in contrast to \texttt{ConstantDistance} which ``walks along the ridge''.
 This discrepancy is even stronger for larger datasets and thus necessitates the use of operators such as \texttt{ConstantDistance} which account for correlations between branch lengths and rates. }
\label{fig:landscape}
\end{figure}

 \texttt{Scale} and \texttt{CisScale} both operate on the clock model standard deviation $\sigma$ however they behave differently in the \textit{real} and \textit{quant} parameterisations (\textbf{Fig \ref{fig:clockSDoperators}}).
In \textit{real}, large proposals of $\sigma \rightarrow \sigma^\prime$ made by  \texttt{Scale} are likely to be rejected due to large penalties in clock model prior $p(\vec{\mathcal{R}}^{\,}|\sigma^\prime)$.
This led to the development of the fast clock scaler \cite{zhang2020improving} (herein referred to as \texttt{CisScale}), which recomputes all branch rates $\vec{\mathcal{R}}^{\,} \rightarrow \vec{\mathcal{R}}^{\, \prime}$ such that their quantiles under the new clock model prior remain constant $p(\vec{\mathcal{R}}^{\,}|\sigma) = p(\vec{\mathcal{R}}^{\, \prime}|\sigma^\prime)$.
In contrast, a proposal made by \texttt{Scale} $\sigma \rightarrow \sigma^\prime$ under the \textit{quant} parameterisation implicitly alters all branch rates $r(\vec{\mathcal{R}}^{\,})$ while leaving the quantiles $\vec{\mathcal{R}}^{\,}$ themselves constant.
But by applying \texttt{CisScale} under \textit{quant},
all quantiles are recomputed $\vec{\mathcal{R}} \rightarrow \vec{\mathcal{R}}^\prime$  such that the respective branch rates are constant i.e. $r(\vec{\mathcal{R}}^{\,}) = r(\vec{\mathcal{R}}^{\,\prime})$.
In summary, \texttt{Scale} and \texttt{CisScale} propose rates/quantiles in the opposite (trans) or same (cis) space that the clock model is parameterised under (\textbf{Fig \ref{fig:clockSDoperators}}).

@@

where \texttt{RF} is the Robinson-Foulds tree distance \cite{robinson1981comparison}, and $|p|$ is the number of dimensions of numerical parameter $p$ (1 for $\sigma$, $2N-2$ for $\vec{\mathcal{R}}^{\,}$, and $2N-1$ for node heights $t$).
The remaining terms are trained during the second and third phases: the cumulative computational runtime spent on each operator $\mathbb{T}(\omega_i)$, the sample standard deviation $\sigma_p$ of each numerical parameter $p$, and the summed distances $\sum_x \mathbb{D}(x_p, x_p^\prime)$.

@@

Datasets with strong signal (or large $L$) are likely to achieve faster convergence when the more precise and meticulous kinds of operators are employed, such as those informed by correlations within the posterior distribution (\textbf{Fig \ref{fig:landscape}}).
Whereas, datasets which contain very poor signal (or small $L$) are likely to mix faster when more weight is placed on bold operators. 
We therefore introduce the \texttt{SampleFromPrior($\vec{x}^{\,}$)} operator.
This operator resamples $\psi$ randomly selected elements within vector $\vec{x}^{\,}$ from their prior distributions, where $\psi \sim \text{Binomial}(n=|\vec{x}^{\,}|, p=\frac{s}{|\vec{x}^{\,}|})$ for tunable term $s$.
%The \texttt{Uniform} operator is a special case of \texttt{SampleFromPrior}.
\texttt{SampleFromPrior} is always included among the mixture of operators under \texttt{AdaptiveOperatorSampler} and serves to make the boldest proposals for datasets with poor signal.
Ideally, an operator to the likes of \texttt{AdaptiveOperatorSampler} would learn the combination of weights behind these classes of operators best suited for any given dataset.

@@

The step size of a proposal kernel $q(x^\prime|x)$ should be such that the proposed state $x^\prime$ is sufficiently far from the current state $x$ to explore vast areas of parameter space, but not so large that the proposal is rejected too often \cite{roberts1997weak}. 
Operators which attain an acceptance probability of 0.234 are often considered to have arrived at a suitable midpoint between these two extremes \cite{bouckaert2019beast, roberts1997weak}.
The standard uniform distribution kernel has recently been challenged by the Bactrian kernel \cite{yang2013searching, thawornwattana2018designing}.
The $\text{Bactrian}(m)$ distribution is defined as the sum of two Normal distributions:

@@

where $0 \leq m < 1$ describes the modality of the Bactrian distribution. When $m=0$, the Bactrian distribution is equivalent to a $\text{Normal}(0, 1)$ distribution. As $m \rightarrow 1$, the distribution becomes increasingly bimodal (\textbf{Fig \ref{fig:bactrian}}). 
Yang et al. 2013 \cite{yang2013searching} suggest that $\text{Bactrian}(m=0.95)$ yields a proposal kernel which traverses the posterior distribution more efficiently that the standard uniform kernel, by placing minimal probability on steps which are too small or too large.
In this case, a target acceptance probability of around 0.3 is optimal.

@@

% Give the equation of the Uniform + Bactrian kernels
% Fig: Bactrian m=0, 0.92, 0.95, 0.98, + Uniform(-1,1)
% In this article we compare the Uniform + Bactrian proposals in the clock model

@@

The \texttt{NarrowExchange} operator \cite{drummond2002estimating}, used widely in BEAST \cite{drummond2012bayesian,suchard2018bayesian} and BEAST2 \cite{bouckaert2019beast}, is similar to nearest-neighbour-interchange \cite{semple2003phylogenetics}, and works as follows (\textbf{Fig \ref{fig:narrowexchange}}):

@@

We hypothesised that if \texttt{NarrowExchange} was adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal (akin to the constant distance operators \cite{zhang2020improving}), then its ability to traverse the state space may improve.

@@

for random walk step size $s\Sigma$ where $s$ is a tunable scaler parameter and $\Sigma$ is drawn from a uniform or \textbf{\nameref{sect:randomwalks}}. NER (and NERw) are compatible with both the \textit{real} and \textit{quant} parameterisations. 
Analogous to the \texttt{ConstantDistance} operator, the proposed rates ensure that the genetic distances between nodes $A$, $B$, $C$, and $E$ are constant. 
There are six pairwise distance between these four nodes and therefore six constraints:

@@

\paragraph{2. Solving Jacobian determinants.} The determinant of the Jacobian matrix $J$ is required for computing the Green ratio of the proposal. $J$ is defined as

@@

\paragraph{3. Automated generation of BEAST2 operators.} Java class files are generated using string processing. Each class corresponds to a single operator, extends the class of a meta-NER-operator, and consists of the solutions found in \textbf{1} and the Jacobian determinant found in \textbf{2}. $|J|$ is further augmented if the \textit{quant} parameterisation is employed (\textbf{\nameref{sect:S1Appendix}}).
Two such operators are expressed in \textbf{Algorithms \ref{alg:NER1} and \ref{alg:NER2}}.

@@

These experiments showed that NER variants which satisfy the genetic distances between nodes $B$ and $A$ (i.e. $\mathcal{D}_{AB}$) or between $B$ and $C$ (i.e. $\mathcal{D}_{BC}$) usually performed worse than the standard \texttt{NarrowExchange} operator (\textbf{Fig \ref{fig:acceptanceRateScreening}}).
This is an intuitive result. 
If there is high uncertainty in the positioning of $B$ with respect to $A$ and $C$, then there is no value in respecting either of these distance constraints, and the proposals made to the rates may often be too bold or the Green ratio $|J|$ too small for the proposal to be accepted.

@@

\subsection*{Round 3: Bactrian proposal kernels are around 15\% more efficient than uniform kernels}
